---
title: "Notebook_pp3_alg_combining_variables"
author: "Armelle Le Guelte"
date: "12/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
```


```{r}
acs1_race <-read_csv(('data/race_Taylor.csv')) 
acs1_sex_age <- read_csv(('data/sex_by_age_Taylor.csv'))
acs1_sex_age2 <- read_csv(('data/sex_by_age.csv')) # need to remove it after
acs5_sex_age_race <- read_csv('data/acs5y2018sex_age_race.csv')
acs5_education <-read_csv(('data/acs5y2018education2.csv'))
```

# Need to put all the data in the same format to use it in the app
# I don't have Name for Metro area and we will use this but I do have metro area numbers
# For a matter of time we decided to pivot the data longer (It was supposed to be wider but it make it a little more complicated to plot the data with ggplot)
# I think to make it work (in wider format) we need to include "pivot_longer" in the formula to plot the data

```{r}
# Make a new dataframe with Metro area and Name
geo_id_info <- acs1_race %>% 
  select(-c("Total", "White alone", "Black or African American alone", "American Indian and Alaska Native alone", "Some other race alone", "Two or more races:", "Two races including Some other race", "Two races excluding Some other race, and three or more races", "Asian alone", "Native Hawaiian and Other Pacific Islander alone", "GEO_ID"))
```

### Cleaning education file

```{r}
# Join education with geo_id_info
acs5_education <-
        inner_join(acs5_education, geo_id_info, by = "metropolitan_statistical_area_micropolitan_statistical_area", copy = FALSE)
```

```{r}
# Change total percentage for 100 because it the estimate number for some weird reason
acs5_education$percent_estimate[acs5_education$percent_estimate > 100] <- 100

```

# I talked to the group and we won't use percentage in the dataframe

```{r}
#Remove percentage to be consistent with Taylor's data
acs5_education <- acs5_education %>% 
  select(-c(percent_estimate))
```


```{r}
# Drop metro area and rename Name as metro_area
acs5_education <- acs5_education %>% 
  filter(education != "Total") %>%
  select(-c(metropolitan_statistical_area_micropolitan_statistical_area)) %>% 
  rename("metro_area" = NAME)
```

```{r}
# Save it as a csv file
write.csv(acs5_education, "data/acs_education_final.csv", row.names = FALSE)
```

### Cleaning race file from Taylor

```{r}
# Remove columns
acs1_race <- acs1_race %>% 
  select(-c("metropolitan_statistical_area_micropolitan_statistical_area", "Total", "GEO_ID"))
```

```{r}
# Rename column
acs1_race <- acs1_race %>% 
  rename("metro_area" = NAME)
```


```{r}
# Pivot the data on race
acs1_race <- pivot_longer(
  acs1_race,
  cols = -"metro_area",
  names_to = "race",
  values_to = "estimate"
)
```

```{r}
# remove the ":" for the line "Two or more races:"
acs1_race$race[acs1_race$race == "Two or more races:"] <- "Two or more races"

```

```{r}
# Save it as a csv file
write.csv(acs1_race, "data/acs_race_final.csv", row.names = FALSE)
```

### Cleaning the sex-age file from Taylor

```{r}
# Remove columns and rename column
acs1_sex_age <- acs1_sex_age %>% 
  select(-c("GEO_ID", "Total")) %>% 
  rename("metro_area" = NAME)
```

```{r}
acs1_sex_age <- pivot_longer(
  acs1_sex_age,
  cols = -"metro_area",
  names_to = "sex_age",
  values_to = "estimate"
)
```


```{r}
# Split the 'label' columns into 8 columns and keep the 'label' column
acs1_sex_age <- acs1_sex_age %>%
     separate(
     sex_age,
     c("sex", "age"),
     sep = ":",
     remove = TRUE,
     convert = FALSE
     )
```

```{r}
# Put "none" in age column for rows without values
acs1_sex_age$age[acs1_sex_age$age == ""] <- "none"
```

```{r}
# Remove total values
acs1_sex_age <- acs1_sex_age %>% 
  filter(age != "none")
```

```{r}
# Put back data wider to be able to sum age columns
acs1_sex_age <- acs1_sex_age %>% 
  pivot_wider(
  id_cols = c("metro_area", "sex"), 
  names_from = "age",
  values_from = "estimate"
)
```

```{r}
# Sum age columns by age categories and remove other columns
acs1_sex_age <- acs1_sex_age %>%
  mutate(`Under 20 years` = `Under 5 years` + `5 to 9 years` + `10 to 14 years` + `15 to 17 years` + `18 and 19 years`,
         `20 to 29 years` = `20 years` + `21 years` + `22 to 24 years` + `25 to 29 years`,
         `30 to 39 years` = `30 to 34 years` + `35 to 39 years`,
         `40 to 49 years` = `40 to 44 years` + `45 to 49 years`,
         `50 to 59 years` = `50 to 54 years` + `55 to 59 years`,
         `60 years and over` = `60 and 61 years` + `62 to 64 years` + `65 and 66 years` + `67 to 69 years` + `70 to 74 years` + `75 to 79 years` + `80 to 84 years` + `85 years and over`
         ) %>% 
select(c(`metro_area`, `sex`, `Under 20 years`, `20 to 29 years`, `30 to 39 years`, `40 to 49 years`, `50 to 59 years`, `60 years and over`))
```

```{r}
# Pivot longer to be consistent with other data for the shiny app
acs1_sex_age <- acs1_sex_age %>% 
  pivot_longer(
    cols = -c("metro_area", "sex"),
    names_to = "age",
    values_to = "estimate"
  )
```


```{r}
# Save it as a csv file
write.csv(acs1_sex_age, "data/sex_by_age.csv", row.names = FALSE)
```

